Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-22 sections/intro.tex
Update on Overleaf.

%
\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
%
This trade-off is
visualized in figure \ref{fig:datavsprogram}.

\begin{table}[h!]
  \help{This table needs citations, as well as a little bit of double checking}
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            & \checkmark \\
%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
  \end{table}

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
  Sparse                   &            & \checkmark &            &            & \checkmark \\
  Run-length               &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
  Ragged                   &            &            & \checkmark &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}

>>>>

%
\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
%
This trade-off is visualized in figure \ref{fig:datavsprogram}.
%
We hypothesize that

\begin{table}[h!]
  \help{This table needs citations, as well as a little bit of double checking}
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &         & \checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
  \end{table}

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmkar & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}
